Multi-redshift limits on the 21cm power spectrum from PAPER

### 2014-10-31 ms.tex
finished reviewer comments, sent to collab

\begin{abstract}
The epoch of reionization power spectrum is expected to evolve strongly with redshift, and it is this variation with cosmic history that will allow us to begin to place constraints on the physics of reionization.  The primary obstacle to the measurement of the EoR power spectrum is bright foreground emission.  We present an analysis of observations from the Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER) telescope which place new limits on the HI power spectrum over the redshift range of $7.5<z<10.5$, extending previously published single redshift results to cover the full range accessible to the instrument.   To suppress foregrounds, we use filtering techniques that take advantage of the large instrumental bandwidth to isolate and suppress foreground leakage into the interesting regions of $k$-space.  Our 500 hour integration is the longest such yet recorded and demonstrates this method to a dynamic range of $10^4$.   Power spectra at different points across the redshift range reveal the variable efficacy of the foreground isolation.  Noise limited measurements of $\Delta^2$ at $k=$0.2hMpc$^{-1}$ and z$=7.55$ reach as low as (48mK)$^2$ ($1\sigma$).  We demonstrate that the size of the error bars in our power spectrum measurement as generated by a bootstrap method is consistent with the fluctuations due to thermal noise.  Relative to this thermal noise, most spectra exhibit and excess of power at a few sigma.  The likely sources of this excess include residual foreground leakage, particularly at the highest redshift, and unflagged RFI.   We conclude by discussing data reduction improvements that promise to remove much of this excess. 
\end{abstract}

@@

Here we report on deep integrations made with a 32 element array in 2011, first described in \cite{Parsons:2014p10499}, hereafter P14.  Our data reduction pipeline was described in detail in P14, where the same methods were used to give the deepest limits yet on the HI power spectrum in the presence of bright foregrounds at redshift 7.68.  However, the Epoch of Reionization signal is expected to evolve strongly with redshift. In fact, it is this signature variation which will distinguish it from foregrounds and it is this variation with cosmic history that will allow us to begin to place constraints on the physics of reionization \citep{Pritchard:2008p8123,Pober:2014p10390}.  Therefore, while a detection of the 21 cm signal at even a single frequency would be a tremendous breakthrough, analysis techniques must be developed to capitalize on the wide bandwidths of the current generation of high-redshift 21cm telescopes.  Using the same data set as P14, this paper presents improved upper limits on the HI power spectrum over the redshift range $10.7>z>7.2$.  In Section  \ref{sec:observations} we summarize the observations, in Section \ref{sec:obs_meth} review the reduction methodology, we present the new upper limits in Section \ref{sec:results}, and in Section \ref{sec:conclusion} we offer conclusions and discussion of future work.

\section{Observations}
\label{sec:observations}
The work here follows the same basic procedure and uses the same underlying data set as P14. Here we provide a quick summary and refer the reader to P14 for a more in-depth discussion.  A general overview of the PAPER system can be found in \cite{Parsons:2010p6757}, calibration of the primary beam in \cite{Pober:2012p8800}, and imaging results in \cite{Jacobs:2011p8438,jacobs:2013b} and \citet{Stefan:2013p9926}.  Sensitivity analysis described in \cite{Parsons:2012p9028} revealed that for the low gain elements employed by PAPER, a highly redundant ``grid'' type arrangement offers a significant sensitivity boost extending the performance of the compact designs motivated by \citet{Morales:2004p2494,Bowman:2006p1887,Lidz:2008p8251}.  In most interferometers the locations of the antennas are optimized such that each baseline samples a different Fourier mode of the sky; this is the ideal case for reconstructing images where each mode contains different information.  For a power spectrum measurement the key metric is sensitivity per mode, rather than number of modes.  A grid configuration allows many samples of each cosmological mode, to be averaged to a high sensitivity before being combined with other Fourier modes.  The PAPER South Africa 32 antenna deployment (PSA32) was arranged in a 4$\times$8 grid with a column spacing of 30m and a row spacing of 4m.  In our analysis, as in P14, we include only the three shortest types of spacings where the reionization power is expected to be brightest. This selection includes those between adjacent columns and within at least one row of each other, a selection containing 70 $\sim$30m-long baselines.  We will use these baselines to make a one dimensional estimate --spectral line of sight only-- of the HI power spectrum.

@@

This isolation of foregrounds to a region below a line-of-sight $k_\parallel$ mode that increases with baseline length is also the much discussed ``wedge'' \citep{PhysRevD.90.023018,PhysRevD.90.023019,Thyagarajan:2013p10039,Pober:2013p9942,Trott:2012p10466,Morales:2012p8790,Parsons:2012p8896,Vedantham:2012p10297,Datta:2010p8781,Parsons:2009p7859}. As we only have three baseline types of nearly the same length the wedge manifests as a single value of $k\sim k_\parallel$, below which foregrounds are expected to dominate.  The details of each application of the delay/fringe-rate transform will be laid out in the following sections as we provide a brief walk-through of the processing pipeline.

@@

\subsection{Calibration}
We model the gain as a per-antenna amplitude and a phase slope -physically a single time delay- and single real, low order polynomial passband for all antennas.  Because the array samples correlations redundantly, the relative calibration between antenna is numerically overdetermined and tractable as a linear algebra problem \citep{Liu:2010p10391}.  As described in P14, we compute the ratio between redundant baselines, fit for a gain and phase slope and then algebraically solve for a per-antenna solution.  Using this method we have avoided calibrating each channel independently to preserve as much frequency variation as possible.   These solutions vary little over the three month observing period, exhibiting less than 1\% r.m.s. variation. A single solution derived for the Dec 7 data set is used for the entire observing run. Time and frequency variation of redundant solutions is explored in general in \citet{Zheng:2014p10467}.

Relative calibration allows the gains of different antennas to be calibrated against one another.  However, to obtain a correctly normalized power spectrum, it is necessary to set a flux scale for which we need a flux measurement of a calibration standard.  For this we form a beamform on a known, bright calibration source.  By itself, the redundant calibration does not contain enough phase information to phase coherently to a sky location, there remain two free phase parameters which cannot be solved by redundancy alone.   We find these by fitting a model of Pictor A, Fornax A, and the Crab Nebula during a time when the sky is dominated by these three sources while marginalizing over the unknown\footnote{Unknown in the sense of a joint uncertainty in source flux and primary beam pattern. Though the fluxes of these sources is in some cases fairly well constrained, fluxes at 150MHz are still fairly uncertain as is the PAPER beam in those directions.} apparent flux ratio between the three sources. With the delays in place we are now able form a beam on Pictor A and (for each channel) set the overall amplitude to the calibration value of 382 (f/150MHz)$^{-0.77}$Jy found in \cite{jacobs:2013b}.

@@

This leaves us with 40 delay samples on each of 70 baselines which are divided into three redundant groups. Within these groups we cross correlate delay spectra between different redundant baselines.  The cross multiplication of the same delay modes between different redundant baselines provides an unbiased estimate of the power spectrum.  These ``sky-like'' correlations should be identical between all redundant baselines to within the level of the noise, while all other cross multiplications between delay modes should not be correlated between different baselines. In practice  these non-sky-like modes do occasionally have significant power which leaks into the correlations which sample the power spectrum.  These are removed by iteratively dividing the covariance into a model of systematics and a model of sky-like emission and then projecting out large residual modes. This is done by dividing the baselines into different groups such that all cross-multiplications are done without introducing noise bias.  For more see Appendix C of P14.

The residual elements of the correlation matrix corresponding to cross-multiplication of matching delay bins between different baselines are all un-biased samples of the power spectrum. To estimate the final power spectrum and its uncertainty we compute the mean and variance of many random randomly-selected subsamples, sampling the dimensions: sidereal time, redundant baseline pair, and delay sign\footnote{As visibilities are complex, both the positive and negative delays  carry  information. Physically the two signs correspond to the two halves of the sky.}. For simplicity, measurements are averaged with equal weights.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/PAPER_32T_new_redshifts.png}
\caption{ \label{fig:pspecs} PAPER power spectra at four redshifts. On the top, $P_k$ spectra (a simple units change from the raw delay spectrum) provide a useful diagnostic on foreground rejection, while on the bottom we plot in $\Delta^2$ cosmological units with 1 $\sigma$ error bars.    Redshift increases left to right. All spectra have an effective bandwidth of $\pm$10MHz  covering the redshift span $\Delta z\approx$1421MHz ($B/f^2$) which ranges from 0.8 at $z=$10.3 to 0.45 at $z=$7.4.  The noise curve (dashes) is calculated using the method described in \cite{Pober:2013p9581}  and indicates the 1$\sigma$ confidence bounds on data points consisting purely of noise; 65\% of uncorrelated noise like data points will lie below the curve. However, note the caveat that due to the weighting of the delay transform, adjacent $k$ bins are 50\% correlated.  See \cite{Pober:2014p10390,Pober:2013p9581} for a discussion of the approximations made in those calculations. The black line is a fiducial model at 50\% ionization \citep{Lidz:2008p8251}.  GMRT points from \cite{Paciga:2013p9943} indicated with 'x's and MWA points (also using 32 antenna) are black diamonds \cite{Dillon:2014p9788}.   }
\end{figure*}

@@

\section{Results}
\label{sec:results}
\subsection{Foreground Filtering and Noise Levels}
\label{sec:noise}
The root-mean-square of differenced visibilities measures how well we've removed foregrounds.  In Figure \ref{fig:noise} we examine $T_{rms}$ as calculated by differencing between adjacent channels and between redundant baselines.  The square of this difference is then averaged over the full LST range probed by the power spectra (1h-9h). Note that this operation combines all LST modes as though they were repeated measurements of the same sky when in practice the power spectrum values are averaged in quadrature over different LST bins. This provides a lower bound on the sensitivity possible in the data set and a sensitive probe on foreground residuals or other sources of excess power.  Using all the data in this way, we expect to illuminate rare outliers that effect the bottom line.

@@

\subsection{The Excess Power}

As these measurements have a maximum possible sensitivity level which is still two orders of magnitude away from theoretical reionization levels we consider the likelihood that these detections are evidence of reionization to be small. Far more likely is that we are limited by a residual foreground or systematic arising from approximations in the data analysis.  The statistical significance of the residual --2$\sigma$ after averaging over the entire season-- is low enough that it is not possible to break it down along likely axes like sidereal time or time of day to probe its origin.  Such studies must await higher sensitivity measurements with more antennas.  However, there are several additional features that seem to suggest an excess not caused by high redshift HI.

The spectral signature of the measured excess is far more similar to a foreground residual then to theoretical HI models.
The predicted signature of the HI signal during reionization is one of a relatively short burst of power at small $k$ values as the ionized bubbles are briefly at their largest, before the signal from the ever increasing ionized medium drops sharply \citep{Pritchard:2008p8123}. In contrast, this observed residual is relatively constant with redshift while rising at both ends of the band, mimicking the RMS noise curve in Figure \ref{fig:noise}.  This suggests that the accuracy of the foreground filter decreases towards the edges of the band.  Recall that the wide-band foreground model we have subtracted was built by weighting the entire band by a Blackman-Harris window function which provides a much higher isolation of foregrounds inside the horizon. The tradeoff that is made is that data in the outer parts of the spectrum are heavily down weighted.  For example data at 126MHz are down-weighted $\sim$75\% compared to 150Mhz.  This amounts to making an assumption about the spectral smoothness of the data, namely that 126MHz can be mostly modeled by extrapolation from 150MHz.  The precision of this subtraction can be judged by examining residuals inside the horizon in the $P(k)$ power spectra plotted in the upper part of Figure \ref{fig:pspecs}\footnote{Note that this notch is not present in Figure \ref{fig:P14compare} because the foregrounds have been left in for comparison with P14.}.  The relative depth of the redshift 10.5 P(k) foreground residual is substantially shallower then at 7.55 and is generally correlated with the amount of residual found at higher $k$s.

The residual level is also illustrated in the $k$ slices of Figure \ref{fig:slices}.  
 The shortest $k$ mode (upper left of Figure \ref{fig:slices}) is the nearest to the light travel horizon and therefore most likely to be contaminated by foreground leakage. This $k$ mode is well above the noise and has a redshift dependence much the same shape as the $T_{rms}$ curve in Figure \ref{fig:noise}, with a minimum near redshift 8.5 and a dramatic rise above redshift 10. These points are also the only ones where error bars significantly deviate from their predicted thermal values. As we see in Table \ref{tab:data}, going by predicted noise alone points closest to the horizon have error bars in excess of 20 $\sigma$ while the bootstrap error bars register a much larger uncertainty.  This hints at a more complicated origin for the excess. If these points were purely foregrounds, the error bars would be noise dominated, the fact that the excess is not much larger then its variation suggests a source of error not accounted in our analysis. Likely culprits include incomplete flagging of interference, instrumental glitches, or time of frequency dependent calibration variation which introduces differences between redundant baselines. All of which have are part of ongoing analysis efforts which are briefly discussed in the conclusions.   All of these effects have the effect of adding time and frequency dependence to the bright foregrounds limiting the effectiveness of the filter.   Thus, just like foregrounds, the strong excess drops rapidly towards higher $k$ modes. Above $k$ of 0.1 hMpc$^{-1}$  the power spectra are generally positively biased, though most close to 2 $\sigma$ or less using either theoretical and bootstrap error bars.

>>>>

\begin{abstract}
The epoch of reionization power spectrum is expected to evolve strongly with redshift, and it is this variation with cosmic history that will allow us to begin to place constraints on the physics of reionization.  The primary obstacle to the measurement of the EoR power spectrum is bright foreground emission.  We present an analysis of observations from the Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER) telescope which place new limits on the HI power spectrum over the redshift range of $7.5<z<10.5$, extending previously published single redshift results to cover the full range accessible to the instrument.   To suppress foregrounds, we use filtering techniques that take advantage of the large instrumental bandwidth to isolate and suppress foreground leakage into the interesting regions of $k$-space.  Our 500 hour integration is the longest such yet recorded and demonstrates this method to a dynamic range of $10^4$.   Power spectra at different points across the redshift range reveal the variable efficacy of the foreground isolation.  Noise limited measurements of $\Delta^2$ at $k=$0.2hMpc$^{-1}$ and z$=7.55$ reach as low as (48mK)$^2$ ($1\sigma$).  We demonstrate that the size of the error bars in our power spectrum measurement as generated by a bootstrap method is consistent with the fluctuations due to thermal noise.  Relative to this thermal noise, most spectra exhibit and excess of power at a few sigma.  The likely sources of this excess include residual foreground leakage, particularly at the highest redshift, unflagged RFI, and calibration errors.   We conclude by discussing data reduction improvements that promise to remove much of this excess. 
\end{abstract}

@@

Here we report on deep integrations made with a 32 element array in 2011, first described in \cite{Parsons:2014p10499}, hereafter P14.  Our data reduction pipeline was described in detail in P14, where the same methods were used to give the deepest limits yet on the HI power spectrum in the presence of bright foregrounds at redshift 7.68.  However, the Epoch of Reionization signal is expected to evolve strongly with redshift. In fact, it is this signature variation which will distinguish it from foregrounds and it is this variation with cosmic history that will allow us to begin to place constraints on the physics of reionization \citep{Pritchard:2008p8123,Pober:2014p10390}.  Therefore, while a detection of the 21 cm signal at even a single frequency would be a tremendous breakthrough, analysis techniques must be developed to capitalize on the wide bandwidths of the current generation of high-redshift 21cm telescopes.

Using the same data set as P14, this paper presents improved upper limits on the HI power spectrum over the redshift range $10.7>z>7.2$.  The power spectrum error bars, calculated both from theory and using a bootstrap, are broadly consistent with a detection of power at the 1$\sigma$ level which can be interpreted as some combination of residual foregrounds, analysis errors and reionization. To determine which components are the mostly likely causes we compare the power spectrum excess measurements of the system noise found by differencing adjacent channels and redundant baselines.

 In Section  \ref{sec:observations} we summarize the observations, in Section \ref{sec:obs_meth} review the reduction methodology, we present the new upper limits in Section \ref{sec:results}, and in Section \ref{sec:conclusion} we offer conclusions and discussion of future work.

\section{Observations}
\label{sec:observations}
The work here follows the same basic procedure and uses the same underlying data set as P14. Here we provide a quick summary and refer the reader to P14 for a more in-depth discussion.  A general overview of the PAPER system can be found in \cite{Parsons:2010p6757}, calibration of the primary beam in \cite{Pober:2012p8800}, and imaging results in \cite{Jacobs:2011p8438,jacobs:2013b} and \citet{Stefan:2013p9926}.  Sensitivity analysis described in \cite{Parsons:2012p9028} revealed that for the low gain elements employed by PAPER, a highly redundant ``grid'' type arrangement offers a significant sensitivity boost extending the performance of the compact designs motivated by \citet{Morales:2004p2494,Bowman:2006p1887,Lidz:2008p8251}.  In most interferometers the locations of the antennas are optimized such that each baseline samples a different Fourier mode of the sky; this is the ideal case for reconstructing images where each mode contains different information.  For a power spectrum measurement the key sensitivity metric is sensitivity per mode, rather than number of modes.  A grid configuration allows many samples of each cosmological mode, to be averaged to a high sensitivity before being combined with other Fourier modes; though at the cost of losing foreground information.

The PAPER South Africa 32 antenna deployment (PSA32) was arranged in a 4$\times$8 grid with a column spacing of 30m and a row spacing of 4m.  In our analysis, as in P14, we include only the three shortest types of spacings where the reionization power is expected to be brightest. This selection includes those between adjacent columns and within at least one row of each other, a selection containing 70 $\sim$30m-long baselines.  We will use these baselines to make a one dimensional estimate --spectral line of sight only-- of the HI power spectrum.

@@

This isolation of foregrounds to a region below a line-of-sight $k_\parallel$ mode that increases with baseline length is also the much discussed ``wedge'' \citep{PhysRevD.90.023018,PhysRevD.90.023019,Thyagarajan:2013p10039,Pober:2013p9942,Trott:2012p10466,Morales:2012p8790,Parsons:2012p8896,Vedantham:2012p10297,Datta:2010p8781,Parsons:2009p7859}. As we only have three baseline types of nearly the same length the wedge manifests as a single value of $k\sim k_\parallel$, below which foregrounds are expected to dominate.

A grid configuration is sufficiently different from a traditional interferometer antenna layout that 
there are many tradeoffs to consider. In this configuration foregrounds cannot be imaged, and therefor can only be filtered or blindly subtracted.  Experimentally, the loss of foreground imaging is offset by an enhanced experimental precision. The grid configuration makes repeated measurements of the \emph{same} modes with \emph{independent} antennas, thus the per-antenna gains are a tractable linear algebra problem that does not depend on a sky model \citep{Liu:2010p10391}. This same redundancy also enables the identification of systematically deviant samples, allows us to avoid noise bias, and provides another axis along which to bootstrap.  These are believed to provide the greatest handle on known systematics.  This solution has a small number of free parameters which must be calibrated to a model of the sky to obtain a power spectrum calibrated to a standard flux scale.  Flux and overall bandpass calibration is the area most impacted by the small number of sampled spatial modes.  In practice, the use of synthesis rotation allows for adequate coverage of the brightest source in the sky which we use as our flux standard (see section \ref{sec:calibration}.

 The details of each application of the delay/fringe-rate transform will be laid out in the following sections as we provide a brief walk-through of the processing pipeline.

@@

\subsection{Calibration}
\label{sec:calibration}
We model the gain as a per-antenna amplitude and a phase slope -physically a single time delay- and single real, low order polynomial passband for all antennas.  Because the array samples correlations redundantly, the relative calibration between antenna is numerically overdetermined and tractable as a linear algebra problem \citep{Liu:2010p10391}.  Importantly this solution does not require a model of the sky or instrument (beyond the assumption of redundancy).  
As described in P14, we compute the ratio between redundant baselines, fit for a gain and phase slope and then algebraically solve for a per-antenna solution.  Using this method we have avoided calibrating each channel independently to preserve as much frequency variation as possible.   These solutions vary little over the three month observing period, exhibiting less than 1\% r.m.s. variation. A single solution derived for the Dec 7 data set is used for the entire observing run. Time and frequency variation of redundant solutions is explored in general in \citet{Zheng:2014p10467}.

Relative calibration allows the gains of different antennas to be calibrated against one another.  However, to obtain a correctly normalized power spectrum, it is necessary to set a flux scale for which we need a flux measurement of a calibration standard.  For this we form a beam on a known, bright calibration source.  By itself, the redundant calibration does not contain enough phase information to phase coherently to a sky location, there remain two free phase parameters which cannot be solved by redundancy alone.   We find these by fitting a model of Pictor A, Fornax A, and the Crab Nebula during a time when the sky is dominated by these three sources while marginalizing over the unknown\footnote{Unknown in the sense of a joint uncertainty in source flux and primary beam pattern. Though the fluxes of these sources is in some cases fairly well constrained, fluxes at 150MHz are still fairly uncertain as is the PAPER beam in those directions.} apparent flux ratio between the three sources. With the delays in place we are now able form a beam on Pictor A and (for each channel) set the overall amplitude to the calibration value of 382 (f/150MHz)$^{-0.77}$Jy found in \cite{jacobs:2013b}.  The accuracy of this step is ultimately limited by the accuracy of the beam-forming on Pictor to get a spectrum.  The principle sources of error include beam modeling error and incomplete uv coverage. Correction for both was explored in detail by \citet{jacobs:2013b} used PAPER data to measure the complete Pictor A track through the primary beam. The light curve matched the beam model well over the full hour angle range, suggesting that the beam model is well understood. When integrated over the full track, the uv coverage was found to be very uniform. The noise weighted average was found to yield a 6\% flux error.  This error in the overall offset of all power spectra is sub-dominant to the observed power spectrum error bars.
%which showed that in measuring Pictor over the full range of hour angles one integrates over many beam pointings and uv-samples.  Comparison of multiple sources with catalog values found on overall standard deviation of \~ 6\%.

@@

This leaves us with 40 delay samples on each of 70 baselines which are divided into three redundant groups.  Each redundant baseline samples the same mode on the sky but with different noise. The cross correlation between baselines gives an estimate of the power spectrum without any additional noise bias. In practice we found that not all cross power spectra agreed as they should were the result dominated by the sky; some instrumental artifacts remained. Though only a small number of modes are directly contaminating a larger number are indirectly affected by the delay spectrum point spread function.

This contamination is removed by iteratively subtracting an average of a bootstrap subset of baselines and then building a model of the residual covariance. This is then divided out of the data covariance.  Within this process the redundant samples are subdivided into four groups such that all cross-multiplications are done between the groups to avoid introducing noise bias into the removed covariance. (For more see Appendix C of P14.)  The estimated cross-power delay spectrum is recorded for each bootstrap.  The final power spectrum and its uncertainty are the mean and variance of randomly selected subsets of these bootstraps. This final bootstrap samples the dimensions: sidereal time, redundant baseline pair, delay sign\footnote{As visibilities are complex, both the positive and negative delays  carry  information. Physically the two signs correspond to the two halves of the sky.}, and mean-sky bootstrap. For simplicity, measurements are averaged with equal weights.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/PAPER_32T_new_redshifts.png}
\caption{ \label{fig:pspecs} PAPER power spectra at four redshifts with error bars derived via bootstrap subsampling. On the top, $P_k$ spectra (a simple units change from the raw delay spectrum) provide a useful diagnostic on foreground rejection, while on the bottom we plot in $\Delta^2$ cosmological units with 1 $\sigma$ error bars.    Redshift increases left to right. All spectra have an effective bandwidth of $\pm$10MHz  covering the redshift span $\Delta z\approx$1421MHz ($B/f^2$) which ranges from 0.8 at $z=$10.3 to 0.45 at $z=$7.4.  The noise curve (dashes) is calculated using the method described in \cite{Pober:2013p9581}  and indicates the 1$\sigma$ confidence bounds on data points consisting purely of noise; 65\% of uncorrelated noise like data points will lie below the curve. However, note the caveat that due to the weighting of the delay transform, adjacent $k$ bins are 50\% correlated.  See \cite{Pober:2014p10390,Pober:2013p9581} for a discussion of the approximations made in those calculations. The black line is a fiducial model at 50\% ionization \citep{Lidz:2008p8251}.  GMRT points from \cite{Paciga:2013p9943} indicated with 'x's and MWA points (also using 32 antenna) are black diamonds \cite{Dillon:2014p9788}.   }
\end{figure*}

@@

\section{Results}
\label{sec:results}
\subsection{Foreground Filtering and Noise Levels}
\label{sec:noise}
The root-mean-square of differenced visibilities measures how well we've removed foregrounds.  In Figure \ref{fig:noise} we examine $T_{rms}$ as calculated by differencing between adjacent channels and between redundant baselines, squaring and averaging in LST.  The channel to channel difference is expected to be dominated by noise and residual foregrounds. The baseline to baseline difference will contain all of these as well as any additional variance due to calibration errors.  These errors will all manifest as an excess rms compared to the theoretical noise level. We can also examine the rms at different sensitivity levels, for example before and after long term averaging. Comparing different integration lengths we can see if its averaging down properly.

@@

\subsection{The Excess Power}
\label{sec:excess power}
As these measurements have a maximum possible sensitivity level which is still two orders of magnitude away from theoretical reionization levels we consider the likelihood that these detections are evidence of reionization to be small. Far more likely is that we are limited by a residual foreground or systematic arising from approximations in the data analysis.  The statistical significance of the residual --2$\sigma$ after averaging over the entire season-- is low enough that it is not possible to break it down along likely axes like sidereal time or time of day to probe its origin.  Such studies must await higher sensitivity measurements with more antennas.  However, there are several additional features that seem to suggest an excess not caused by high redshift HI.

The spectral signature of the measured excess is far more similar to a foreground residual then to theoretical HI models.
The predicted signature of the HI signal during reionization is one of a relatively short burst of power at small $k$ values as the ionized bubbles are briefly at their largest, before the signal from the ever increasing ionized medium drops sharply \citep{Pritchard:2008p8123}. In contrast, this observed residual is relatively constant with redshift while rising at both ends of the band, mimicking the RMS noise curve in Figure \ref{fig:noise}.  This suggests that the accuracy of the foreground filter decreases towards the edges of the band.  Recall that the wide-band foreground model we have subtracted was built by weighting the entire band by a Blackman-Harris window function which provides a much higher isolation of foregrounds inside the horizon. The tradeoff is that data in the outer parts of the spectrum are heavily down weighted.  For, in the subtracted foreground model, data at 126MHz are down-weighted $\sim$75\% compared to 150Mhz.  This amounts to making an assumption about the spectral evolution of the data, namely that 126MHz can be mostly modeled by extrapolation from 150MHz.  The precision of this subtraction can be judged by examining residuals inside the horizon in the $P(k)$ power spectra plotted in the upper part of Figure \ref{fig:pspecs}\footnote{Note that this notch is not present in Figure \ref{fig:P14compare} because the foregrounds have been left in for comparison with P14.}.  The relative depth of the redshift 10.5 P(k) foreground residual is substantially shallower then at 7.55 and is generally correlated with the amount of residual found at higher $k$s.

The residual level is also illustrated in the $k$ slices of Figure \ref{fig:slices}.  
 The shortest $k$ mode (upper left of Figure \ref{fig:slices}) is the nearest to the light travel horizon and therefore most likely to be contaminated by foreground leakage. This $k$ mode is well above the noise and has a redshift dependence much the same shape as the $T_{rms}$ curve in Figure \ref{fig:noise}, with a minimum near redshift 8.5 and a dramatic rise above redshift 10. These points are also the only ones where error bars significantly deviate from their predicted thermal values. As we see in Table \ref{tab:data}, going by predicted noise alone points closest to the horizon have error bars in excess of 20 $\sigma$ while the bootstrap error bars register a much larger uncertainty.  This hints at a more complicated origin for the excess. If these points were purely foregrounds, the error bars would be noise dominated, the fact that the excess is not much larger then its variation suggests a source of error not accounted in our analysis. Likely culprits include incomplete flagging of interference, instrumental glitches, time and frequency dependent calibration variation which introduces differences between redundant baselines, and imperfect instrumental redundancy (e.g. variation of primary beams between antennas or deviations from redundant antenna placement). All of which have are part of ongoing analysis efforts which are briefly discussed in the conclusions.   Most of these effects have the effect of adding time and frequency dependence to the bright foregrounds limiting the effectiveness of the filter.   Thus, just like foregrounds, the strong excess drops rapidly towards higher $k$ modes. Above $k$ of 0.1 hMpc$^{-1}$  the power spectra are generally positively biased, though most close to 1 $\sigma$.
